Comparison to Christensen et al. 2018

### 2019-07-10 doc/paper-comparison.Rmd

---
title: "Comparison to Christensen et al. 2018"
author: "Renata Diaz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paper-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

This document prodives a side-by-side comparison of LDATS (v0.1.0) results with analysis from [Christensen et al. 2018](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2373).

@@

@@

To obtain the most recent version of **LDATS**, install the most recent
version from GitHub:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("weecology/LDATS")
```

```{r, eval = T}
library(LDATS)
library(dplyr)
load(here::here('vignettes', 'christensenetal-comparison-files',  'data_files.Rdata'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache',  'ldas.RData'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_changepoints.RData'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'paper_changepoints.RData'))
```

## Christensen 2018 analysis files

Download Christensen 2018 analysis scripts & data files from [Extreme-events-LDA repo:](https://github.com/emchristensen/Extreme-events-LDA)

@@

  * rodent_LDA_analysis.R main script for analyzing rodent community change using LDA

  * rodent_data_for_LDA.R contains a function that creates the rodent data table used in analyses

  * AIC_model_selection.R contains functions for calculating AIC for different candidate LDA models

  * changepointmodel.r contains change-point model code

  * LDA-distance.R function for computing Hellinger distance analyses

@@

  * Rodent_table_dat.csv table of rodent data, created by rodent_data_for_LDA.R

@@

  * LDA_figure_scripts.R contains functions for making main plots in manuscript (Fig 1). Called from rodent_LDA_analysis.R

```{r download Christensen, eval = F}

paper_filepath <- here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files')
test_filepath <- here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'rodent_LDA_analysis.r')

if(!file.exists(test_filepath)) {

files_to_download <- c('rodent_LDA_analysis.r', 'rodent_data_for_LDA.r', 'AIC_model_selection.R', 'changepointmodel.r', 'LDA-distance.R', 'Rodent_table_dat.csv', 'LDA_figure_scripts.R')

for(i in 1:length(files_to_download)) {
  download.file(url = paste0("https://raw.githubusercontent.com/emchristensen/Extreme-events-LDA/master/", files_to_download[i]),
                destfile = paste0(paper_filepath, '/', files_to_download[i]))
}

rm(files_to_download)
rm(i)

}

rm(test_filepath)
rm(paper_filepath)
```

## Data

The Portal rodents control data is included in the LDATS package:

@@

Load the data used in Christensen et al:

```{r Paper data, eval = F}
source(here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'rodent_data_for_LDA.r'))

dat = create_rodent_table(period_first = 1,
                          period_last = 436,
                          selected_plots = c(2,4,8,11,12,14,17,22),
                          selected_species = c('BA','DM','DO','DS','NA','OL','OT','PB','PE','PF','PH','PI','PL','PM','PP','RF','RM','RO','SF','SH','SO'))

# dates to go with count data
moondat = read.csv(text=getURL("https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/moon_dates.csv"),stringsAsFactors = F)
moondat$date = as.Date(moondat$censusdate)

period_dates = filter(moondat,period %in% rownames(dat)) %>% select(period,date)
dates = period_dates$date

paper_dat <- dat

paper_dates <- dates

rm(list = c('dat', 'dates', 'period_dates',
            'create_rodent_table'))

```

Compare paper data to LDATS data:

```{r rodent data comparison}

compare <- rodents[[1]] == paper_dat

@@

There are 16 rows where the data included in LDATS differs from the paper data. This is because the LDATS data is not adjusted to account for trapping effort, but the paper data has divided all census counts by the actual number of plots trapped and multiplied by 8 to account for incompletely-trapped censuses.

@@

```
# retrieve data on number of plots trapped per month
  trap_table = read.csv('https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_trapping.csv')
  trap_table_controls = filter(trap_table, plot %in% selected_plots)
  nplots_controls = aggregate(trap_table_controls$sampled,by=list(period = trap_table_controls$period),FUN=sum)

  # adjust species counts by number of plots trapped that month
  r_table_adjusted = as.data.frame.matrix(r_table)
  for (n in 1:436) {
    #divide by number of control plots actually trapped (should be 8) and multiply by 8 to estimate captures as if all plots were trapped
    r_table_adjusted[n,] = round(r_table_adjusted[n,]/nplots_controls$x[n]*8)
  }

    return(r_table_adjusted)

```

Running the same procedure on the LDATS data, we modify 16 rows, and finish with a data frame that matches the one from the paper.

```{r adjust LDATS data after Christensen et al, eval = F}

trap_table = read.csv('https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_trapping.csv')

```

```{r continue adjust LDATS after paper}
 trap_table_controls =dplyr::filter(trap_table, plot %in% c(2,4,8,11,12,14,17,22))
  nplots_controls = aggregate(trap_table_controls$sampled,by=list(period = trap_table_controls$period),FUN=sum)
   # adjust species counts by number of plots trapped that month
  ldats_rodents_adjusted = as.data.frame.matrix(rodents[[1]])
  for (n in 1:436) {
    #divide by number of control plots actually trapped (should be 8) and multiply by 8 to estimate captures as if all plots were trapped
    ldats_rodents_adjusted[n,] = round(ldats_rodents_adjusted[n,]/nplots_controls$x[n]*8)
  }

compare_raw <- rodents[[1]] == ldats_rodents_adjusted
length(which(rowSums(compare_raw) < ncol(compare_raw)))

compare_adjusted <- ldats_rodents_adjusted == paper_dat
length(which(rowSums(compare_adjusted) < ncol(compare_adjusted)))

```

```{r data adjustment cleanup, include = FALSE}
rm(list = c('compare', 'compare_adjusted', 'compare_raw', 'n', 'ldats_rodents_adjusted', 'nplots_controls', 'trap_table', 'trap_table_controls'))
```

@@

```{r add dates to covariate table}

head(rodents$document_covariate_table)

moondat <- dplyr::select(moondat, newmoonnumber, censusdate)
colnames(moondat) <- c('newmoon', 'censusdate')

new_cov_table <- dplyr::left_join(rodents$document_covariate_table, moondat, by = 'newmoon')

rodents$document_covariate_table <- new_cov_table

```

```{r cleanup dates, include = F}
rm(moondat)
rm(new_cov_table)
```

## Stage 1: LDA

@@

```{r LDATS LDAs, eval = F}

ldats_ldas <- LDATS::LDA_set(document_term_table = rodents$document_term_table, topics = c(2:6), nseeds = 100)
ldats_lda_selected <- LDATS::select_LDA(LDA_models = ldats_ldas)

save(ldats_ldas, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_ldas.Rds'))
save(ldats_lda_selected, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_lda_selected.Rds'))

```

Paper LDAS:

```{r create dat for paper lda, include = F}
dat = paper_dat
```

```{r paper LDAs, eval = F}
source(here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'AIC_model_selection.R'))
source(here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'LDA-distance.R'))

# Fit a bunch of LDA models with different seeds
# Only use even numbers for seeds because consecutive seeds give identical results
seeds = 2*seq(200)

# repeat LDA model fit and AIC calculation with a bunch of different seeds to test robustness of the analysis
best_ntopic = repeat_VEM(paper_dat,
                         seeds,
                         topic_min=2,
                         topic_max=6)
hist(best_ntopic$k,breaks=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5,7.5,8.5,9.5),xlab='best # of topics', main='')

# 2b. how different is species composition of 4 community-types when LDA is run with different seeds?
# ==================================================================
# get the best 100 seeds where 4 topics was the best LDA model
seeds_4topics = best_ntopic %>%
  filter(k == 4) %>%
  arrange(aic) %>%
  head(100) %>%
  pull(SEED)

# choose seed with highest log likelihood for all following analyses
#    (also produces plot of community composition for 'best' run compared to 'worst')
best_seed = calculate_LDA_distance(paper_dat,seeds_4topics)
mean_dist = unlist(best_seed)[2]
max_dist = unlist(best_seed)[3]

# ==================================================================
# 3. run LDA model
# ==================================================================
ntopics = 4
SEED = unlist(best_seed)[1]  # For the paper, use seed 206
ldamodel = LDA(paper_dat,ntopics, control = list(seed = SEED),method='VEM')

save(ldamodel,file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'paper_ldamodel.Rds'))
```

### Plot LDAS

#### Paper LDA
```{r plot paper LDA, fig.width=7, fig.height=6}
# Paper
plot(ldamodel, cols = NULL, option = "D")

```

#### LDATS LDA
```{r plot LDATS LDA, fig.width=7, fig.height=6}
# LDATS
plot(ldats_lda_selected[[1]], cols = NULL, option = "D")
```

The paper method finds 4 topics and LDATS finds 6. This is because of an update to the model selection procedure. The paper conservatively overestimates the number of parameters (by counting all of the variational parameteres) and therefore overpenalizes the AIC for models with more topics. Comparatively, the LDATS method now uses the number of parameters remaining after the variational approximation, as returned by the LDA object. For this vignette, we will compare the results from using both LDA models.

## Changepoint models

@@

### Running paper changepoint models

#### LDATS LDA and paper changepoint

```{r run LDATS LDA and paper cpt, eval = F}
#### Run changepoint ####
source(here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'changepointmodel.r'))

# set up parameters for model
year_continuous = 1970 + as.integer(julian(paper_dates)) / 365.25
x = data.frame(
  year_continuous = year_continuous,
  sin_year = sin(year_continuous * 2 * pi),
  cos_year = cos(year_continuous * 2 * pi)
)

# run models with 1, 2, 3, 4, 5, 6 changepoints
ldats_paper_cp_results_rodent = changepoint_model(ldats_lda_selected[[1]], x, 1, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_1.Rds'))
rm(ldats_paper_cp_results_rodent)

ldats_paper_cp_results_rodent2 = changepoint_model(ldats_lda_selected[[1]], x, 2, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent2, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_2.Rds'))
rm(ldats_paper_cp_results_rodent2)

ldats_paper_cp_results_rodent3 = changepoint_model(ldats_lda_selected[[1]], x, 3, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent3, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_3.Rds'))
rm(ldats_paper_cp_results_rodent3)

ldats_paper_cp_results_rodent4 = changepoint_model(ldats_lda_selected[[1]], x, 4, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent4, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_4.Rds'))
rm(ldats_paper_cp_results_rodent4)

ldats_paper_cp_results_rodent5 = changepoint_model(ldats_lda_selected[[1]], x, 5, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent5, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_5.Rds'))
rm(ldats_paper_cp_results_rodent5)

ldats_paper_cp_results_rodent6 = changepoint_model(ldats_lda_selected[[1]], x, 6, weights = rep(1,length(year_continuous)))
save(ldats_paper_cp_results_rodent6, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_6.Rds'))
rm(ldats_paper_cp_results_rodent6)
```

```{r select LDATS LDA and paper cpt, eval = F}
#### Changepoint model selection ####
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_1.Rds'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_2.Rds'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_3.Rds'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_4.Rds'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_5.Rds'))
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'ldats_paper_6.Rds'))

```

```{r select LDATS-paper, eval = T}
ntopics = ldats_lda_selected[[1]]@k

# change point model selection
# mean deviance ( -2 * log likelihood) + 2*(#parameters)
mean(ldats_paper_cp_results_rodent$saved_lls * -2) + 2*(3*(ntopics-1)*(1+1)+(1))
mean(ldats_paper_cp_results_rodent2$saved_lls * -2)+ 2*(3*(ntopics-1)*(2+1)+(2))
mean(ldats_paper_cp_results_rodent3$saved_lls * -2)+ 2*(3*(ntopics-1)*(3+1)+(3))
mean(ldats_paper_cp_results_rodent4$saved_lls * -2)+ 2*(3*(ntopics-1)*(4+1)+(4))
mean(ldats_paper_cp_results_rodent5$saved_lls * -2)+ 2*(3*(ntopics-1)*(5+1)+(5))
mean(ldats_paper_cp_results_rodent6$saved_lls * -2)+ 2*(3*(ntopics-1)*(6+1)+(6))

# The lowest deviance is for 4 changepoints.

ldats_paper_cpt_selected = ldats_paper_cp_results_rodent4

df_4 = as.data.frame(t(ldats_paper_cpt_selected$saved[,1,])) %>% reshape::melt()
year_continuous = 1970 + as.integer(julian(paper_dates)) / 365.25
df_4$value = year_continuous[df_4$value]

ldats_paper_cpt_dates <- vector(length = 4)
ldats_paper_cpt_dates[1] <- mean(df_4[df_4$variable=='V1','value']) %>% lubridate::date_decimal() %>% format('%Y-%m-%d')
ldats_paper_cpt_dates[2] <-mean(df_4[df_4$variable=='V2','value']) %>% lubridate::date_decimal() %>% format('%Y-%m-%d')
ldats_paper_cpt_dates[3] <-mean(df_4[df_4$variable=='V3','value']) %>% lubridate::date_decimal() %>% format('%Y-%m-%d')
ldats_paper_cpt_dates[4] <-mean(df_4[df_4$variable=='V4','value']) %>% lubridate::date_decimal() %>% format('%Y-%m-%d')

ldats_paper_cpt_dates

```

#### Paper LDA and paper changepoint

```{r load paper LDA for paper cpt, eval = F}

#### Load LDA ####
load(here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'paper_ldamodel.Rds'))

```

```{r run paper LDA and paper cpt, eval = F}
#### Run changepoint ####
source(here::here('vignettes', 'christensenetal-comparison-files', 'christensenetal-2018-files', 'changepointmodel.r'))

# set up parameters for model
year_continuous = 1970 + as.integer(julian(paper_dates)) / 365.25
x = data.frame(
  year_continuous = year_continuous,
  sin_year = sin(year_continuous * 2 * pi),
  cos_year = cos(year_continuous * 2 * pi)
)

# run models with 1, 2, 3, 4, 5, 6 changepoints
paper_paper_cp_results_rodent = changepoint_model(ldamodel, x, 1, weights = rep(1,length(year_continuous)))
save(paper_paper_cp_results_rodent, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'paper_paper_1.Rds'))
rm(paper_paper_cp_results_rodent)

paper_paper_cp_results_rodent2 = changepoint_model(ldamodel, x, 2, weights = rep(1,length(year_continuous)))
save(paper_paper_cp_results_rodent2, file = here::here('vignettes', 'christensenetal-comparison-files', 'model-cache', 'paper_paper_2.Rds'))
rm(paper_paper_cp_results_rodent2)

>>>>

---
title: "Comparison to Christensen et al. 2018"
author: "Renata Diaz and Hao Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paper-comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  run_models: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=FALSE}
vers <- packageVersion("LDATS")
```

# Introduction

This document provides a side-by-side comparison of **`LDATS`** (version `r vers`) results with analysis from [Christensen et al. 2018](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2373).

@@

# Setup

@@

To obtain the most recent version of **`LDATS`**, install the most recent version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("weecology/LDATS")
```

Load in the **`LDATS`** package.
```{r}
library(LDATS)
set.seed(42)
nseeds <- 200
nit <- 10000
```

## Running the Models

Because both the Latent Dirichlet Allocation (LDA) and time series components of the analysis can take a long time to run (especially with the settings above for the number of seeds and iterations), we will use pre-generated model outputs and turn off certain code chunks that run the models using a global `rmarkdown` parameter, `run_models = FALSE`.

To change this functionality, you can re-render this file with:
```{r, eval = FALSE}
rmarkdown::render("paper-comparison.Rmd", params = list(run_models = TRUE))
```

## Download Analysis Scripts and Data Files

We're going to download analysis scripts, data files, and model objects, so we use a temporary location for storage:

```{r set download location}
vignette_files <- tempdir()
```

To replicate the Christensen et al. 2018 analysis, we download some of the original scripts & data files from [Extreme-events-LDA repo](https://github.com/emchristensen/Extreme-events-LDA), as well as some raw data files from the [PortalData repo](https://github.com/weecology/PortalData):

@@

* `rodent_LDA_analysis.R`
  - main script for analyzing rodent community change using LDA

* `rodent_data_for_LDA.R`
  - contains a function that creates the rodent data table used in analyses

* `AIC_model_selection.R`
  - contains functions for calculating AIC for different candidate LDA models

* `changepointmodel.r`
  - contains change-point model code

* `LDA-distance.R`
  - function for computing Hellinger distance analyses

@@

* `Rodent_table_dat.csv`
  - table of rodent data, created by rodent_data_for_LDA.R

* `moon_dates.csv`
  - table of census dates (downloaded from the PortalData repository)

* `Portal_rodent_trapping.csv`
  - table of trapping effort (downloaded from the PortalData repository)

@@

* `LDA_figure_scripts.R`
- contains functions for making main plots in manuscript (Fig 1). Called from rodent_LDA_analysis.R

```{r download scripts}
test_file <- file.path(vignette_files, "rodent_LDA_analysis.r")

if (!file.exists(test_file))
{
  # from the Extreme-events-LDA repo
  github_path <- "https://raw.githubusercontent.com/emchristensen/Extreme-events-LDA/master/"
  files_to_download <- c("rodent_LDA_analysis.r", "rodent_data_for_LDA.r", 
                         "AIC_model_selection.R", "changepointmodel.r", 
                         "LDA-distance.R", "Rodent_table_dat.csv", 
                         "LDA_figure_scripts.R")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file))
  }

  # from the PortalData repo
  github_path <- "https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/"
  files_to_download <- c("moon_dates.csv", "Portal_rodent_trapping.csv")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file))
  }
}
```

## Download Model Outputs

We also have pre-generated model outputs that we download from the [LDATS-replications repo](https://github.com/weecology/LDATS-replications):

LDA models:

* `ldats_ldamodel.RDS`
  - the best LDA model as selected by LDATS

* `paper_ldamodel.RDS`
  - the best LDA model as selected by the Christensen et al. analysis

Changepoint outputs

* `ldats_ldats.RDS`
  - the posterior distribution of changepoints, using the LDATS LDA model and the LDATS changepoint selection

* `ldats_paper.RDS`
  - the posterior distribution of changepoints, using the LDATS LDA model and the paper's changepoint selection

* `paper_ldats.RDS`
  - the posterior distribution of changepoints, using the paper LDA model and the LDATS changepoint selection

* `paper_paper.RDS`
  - the posterior distribution of changepoints, using the paper LDA model and the paper's changepoint selection

Figures

* `lda_distances.png`
  - figure showing the variance in the topics identified by the paper's LDA model code

```{r download pre-generated model outputs, eval = !params$run_model}
test_file <- file.path(vignette_files, "ldats_ldamodel.RDS")

if (!file.exists(test_file))
{
  # from the Extreme-events-LDA repo
  github_path <- "https://raw.githubusercontent.com/weecology/LDATS-replications/master/model_outputs/"
  files_to_download <- c("ldats_ldamodel.RDS", "paper_ldamodel.RDS", 
                         "ldats_ldats.RDS", "ldats_paper.RDS", 
                         "paper_ldats.RDS", "paper_paper.RDS", 
                         "lda_distances.png")

  for (file in files_to_download)
  {
    download.file(url = paste0(github_path, file),
                  destfile = file.path(vignette_files, file), 
                  mode = "wb")
  }
}
```

# Data Comparison

The dataset of Portal rodents on control plots is included in the LDATS package:

@@

We can compare this against the data used in Christensen et al:

```{r Paper data}
# parameters for subsetting the full Portal rodents data
periods <- 1:436
control_plots <- c(2, 4, 8, 11, 12, 14, 17, 22)
species_list <- c("BA", "DM", "DO", "DS", "NA", "OL", "OT", "PB", "PE", "PF", 
                  "PH", "PI", "PL", "PM", "PP", "RF", "RM", "RO", "SF", "SH", "SO")

source(file.path(vignette_files, "rodent_data_for_LDA.r"))

# assemble `paper_dat`, the data from Christensen et al. 2018
paper_dat <- create_rodent_table(period_first = min(periods),
                                 period_last = max(periods),
                                 selected_plots = control_plots,
                                 selected_species = species_list)

# assemble `paper_covariates`, the associated dates and covariate data
moondat <- read.csv(file.path(vignette_files, "moon_dates.csv"), stringsAsFactors = F)

paper_dates <- moondat %>%
  dplyr::filter(period %>% dplyr::between(min(periods), max(periods))) %>% 
  dplyr::pull(censusdate) %>%
  as.Date()

paper_covariates <- data.frame(
  index = seq_along(paper_dates), 
  date = paper_dates, 
  year_continuous = lubridate::decimal_date(paper_dates)) %>%
  dplyr::mutate( 
    sin_year = sin(year_continuous * 2 * pi), 
    cos_year = cos(year_continuous * 2 * pi)
  )
```

## Compare the data from Christensen et al. with the included data in `LDATS`

```{r rodent data comparison}
compare <- rodents[[1]] == paper_dat

@@

There are 16 rows where the data included in LDATS differs from the paper data. This is because the LDATS data is not adjusted to account for trapping effort, while the paper data does, by dividing all census counts by the actual number of plots trapped and multiplying by 8 to account for incompletely-trapped censuses.

@@

```{}
  # retrieve data on number of plots trapped per month
  trap_table = read.csv('https://raw.githubusercontent.com/weecology/PortalData/master/Rodents/Portal_rodent_trapping.csv')
  trap_table_controls = filter(trap_table, plot %in% selected_plots)
  nplots_controls = aggregate(trap_table_controls$sampled,by=list(period = trap_table_controls$period),FUN=sum)

  # adjust species counts by number of plots trapped that month
  r_table_adjusted = as.data.frame.matrix(r_table)
  for (n in 1:436) {
    #divide by number of control plots actually trapped (should be 8) and multiply by 8 to estimate captures as if all plots were trapped
    r_table_adjusted[n,] = round(r_table_adjusted[n,]/nplots_controls$x[n]*8)
  }
```

We can run the same procedure on the LDATS data to verify that we obtain a data.frame that matches.

```{r adjust LDATS data after Christensen et al, eval = TRUE}
# get the trapping effort for each sample
trap_table <- read.csv(file.path(vignette_files, "Portal_rodent_trapping.csv"))
trap_table_controls <- dplyr::filter(trap_table, plot %in% control_plots)
nplots_controls <- aggregate(trap_table_controls$sampled, 
                            by = list(period = trap_table_controls$period), 
                            FUN = sum)

# adjust species counts by number of plots trapped that month
#   divide by number of control plots actually trapped (should be 8) and 
#   multiply by 8 to estimate captures as if all plots were trapped
ldats_rodents_adjusted <- as.data.frame.matrix(rodents[[1]])
ldats_rodents_adjusted[periods, ] <- round(ldats_rodents_adjusted[periods, ] / nplots_controls$x[periods] * 8)
```

Now we can compare the adjusted ldats dataset with both the original ldats dataset and the dataset from the paper:

```{r dataset comparisons}
compare_raw <- rodents[[1]] == ldats_rodents_adjusted
length(which(rowSums(compare_raw) < ncol(compare_raw)))

compare_adjusted <- ldats_rodents_adjusted == paper_dat
length(which(rowSums(compare_adjusted) < ncol(compare_adjusted)))
```

@@

```{r add dates to covariate table}
head(rodents$document_covariate_table)

new_cov_table <- dplyr::left_join(rodents$document_covariate_table, 
                                  dplyr::select(moondat, newmoonnumber, censusdate), 
                                  by = c("newmoon" = "newmoonnumber")) %>%
  dplyr::rename(date = censusdate)

rodents$document_covariate_table <- new_cov_table
```

# Identify community groups using LDA

@@

First, we run the LDA models from LDATS to identify the number of topics:

```{r LDATS LDAs, eval = params$run_models}
ldats_ldas <- LDATS::LDA_set(document_term_table = rodents$document_term_table, 
                             topics = 2:6, nseeds = nseeds)
ldats_ldamodel <- LDATS::select_LDA(LDA_models = ldats_ldas)[[1]]

saveRDS(ldats_ldamodel, file = file.path(vignette_files, "ldats_ldamodel.RDS"))
```

Second, we run the LDA models from Christensen et al. to do the same task:

```{r paper LDAs, eval = params$run_models}
source(file.path(vignette_files, "AIC_model_selection.R"))
source(file.path(vignette_files, "LDA-distance.R"))

# Some of the functions require the data to be stored in the `dat` variable
dat <- paper_dat

# Fit a bunch of LDA models with different seeds
# Only use even numbers for seeds because consecutive seeds give identical results
seeds <- 2 * seq(nseeds)

# repeat LDA model fit and AIC calculation with a bunch of different seeds to test robustness of the analysis
best_ntopic <- repeat_VEM(paper_dat,
                          seeds,
                          topic_min = 2,
                          topic_max = 6)
hist(best_ntopic$k, breaks = seq(from = 0.5, to = 9.5), 
     xlab = "best # of topics", main = "")

# 2b. how different is species composition of 4 community-types when LDA is run with different seeds?
# ==================================================================
# get the best 100 seeds where 4 topics was the best LDA model
seeds_4topics <- best_ntopic %>%
  filter(k == 4) %>%
  arrange(aic) %>%
  head(min(100, nseeds)) %>%
  pull(SEED)

# choose seed with highest log likelihood for all following analyses
#    (also produces plot of community composition for "best" run compared to "worst")

png(file.path(vignette_files, "lda_distances.png"), width = 800, height = 400)
dat <- paper_dat # calculate_LDA_distance has some required named variables
best_seed <- calculate_LDA_distance(paper_dat, seeds_4topics)
dev.off()
mean_dist <- unlist(best_seed)[2]
max_dist <- unlist(best_seed)[3]

# ==================================================================
# 3. run LDA model
# ==================================================================
ntopics <- 4
SEED <- unlist(best_seed)[1]  # For the paper, use seed 206
ldamodel <- LDA(paper_dat, ntopics, control = list(seed = SEED), method = "VEM")

saveRDS(ldamodel, file = file.path(vignette_files, "paper_ldamodel.RDS"))
```

```{r}
knitr::include_graphics(file.path(vignette_files, "lda_distances.png"))
```

## Plots

To visualize the LDA assignment of species to topics, we load in the saved LDA models from previously:
```{r}
ldamodel <- readRDS(file.path(vignette_files, "paper_ldamodel.RDS"))
ldats_ldamodel <- readRDS(file.path(vignette_files, "ldats_ldamodel.RDS"))
```

How the paper LDA model assigns species to topics:
```{r plot paper LDA, fig.width = 7, fig.height = 6}
plot(ldamodel, cols = NULL, option = "D")
```

How the LDATS LDA model assigns species to topics:
```{r plot LDATS LDA, fig.width = 7, fig.height = 6}
plot(ldats_ldamodel, cols = NULL, option = "D")
```

The paper method finds 4 topics and LDATS finds 6. This is because of an update to the model selection procedure. The paper conservatively overestimates the number of parameters (by counting all of the variational parameters) and therefore overpenalizes the AIC for models with more topics. Comparatively, the LDATS method now uses the number of parameters remaining after the variational approximation, as returned by the LDA object. For this vignette, we will compare the results from using both LDA models.

# Changepoint models

@@

## Running paper changepoint models

We define a few helper functions for running the changepoints model of Christensen et al. and processing the output to obtain the dates:

```{r paper changepoint models}
#### Run changepoint ####
source(file.path(vignette_files, "changepointmodel.r"))

find_changepoints <- function(lda_model, paper_covariates, n_changepoints = 1:6)
{
  # set up parameters for model
  x <- dplyr::select(paper_covariates, 
                     year_continuous, 
                     sin_year, 
                     cos_year)

  # run models with 1, 2, 3, 4, 5, 6 changepoints
  cpt_results <- data.frame(n_changepoints = n_changepoints)
  cpt_results$cpt_model <- lapply(cpt_results$n_changepoints,
                                  function(n_changepoints)
                                  {
                                    changepoint_model(lda_model, x, n_changepoints, maxit = nit, 
                                                      weights = rep(1, NROW(x)))
                                  })
  return(cpt_results)
}

# Among a selection of models with different # of changepoints, 
#   - compute AIC
#   - select the model with the best AIC
#   - get the posterior distributions for the changepoints
select_cpt_model <- function(cpt_results, ntopics)
{
  # compute log likelihood as the mean deviance
  cpt_results$mean_deviances <- vapply(cpt_results$cpt_model, 
                                       function(cpt_model) {mean(cpt_model$saved_lls)}, 
                                       0)

  # compute AIC = ( -2 * log likelihood) + 2 * (#parameters)
  cpt_results$AIC <- cpt_results$mean_deviances * -2 + 
    2 * (3 * (ntopics - 1) * (cpt_results$n_changepoints + 1) +
           (cpt_results$n_changepoints))

  # select the best model
  cpt <- cpt_results$cpt_model[[which.min(cpt_results$AIC)]]
  return(cpt)
}

# transform the output from `compute_cpt` and match up the time indices with 
#   dates from the original data
get_dates <- function(cpt, covariates = paper_covariates)
{
  cpt$saved[,1,] %>%
    t() %>%
    as.data.frame() %>%
    reshape::melt() %>%
    dplyr::left_join(covariates, by = c("value" = "index"))
}
```

### LDATS LDA and paper changepoint

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by LDATS:
```{r run LDATS LDA and paper cpt, eval = params$run_models}
ldats_paper_results <- find_changepoints(ldats_ldamodel, paper_covariates)

saveRDS(ldats_paper_results, file = file.path(vignette_files, "ldats_paper.RDS"))
```

Extract the dates of the changepoints:
```{r compute changepoints for LDATS LDA and paper cpt}
ldats_paper_results <- readRDS(file.path(vignette_files, "ldats_paper.RDS"))

ldats_paper_cpt <- select_cpt_model(ldats_paper_results, 
                                    ntopics = ldats_ldamodel@k)
ldats_paper_cpt_dates <- get_dates(ldats_paper_cpt)
```

### Paper LDA and paper changepoint

Run the Christensen et al. time series model to identify changepoints on the LDA topics selected by Christensen et al.:
```{r run paper LDA and paper cpt, eval = params$run_models}
paper_paper_results <- find_changepoints(ldamodel, paper_covariates)

saveRDS(paper_paper_results, file = file.path(vignette_files, "paper_paper.RDS"))
```

Extract the dates of the changepoints:
```{r compute changepoints for paper LDA and paper cpt}
paper_paper_results <- readRDS(file.path(vignette_files, "paper_paper.RDS"))

paper_paper_cpt <- select_cpt_model(paper_paper_results, 
                                    ntopics = ldamodel@k)
paper_paper_cpt_dates <- get_dates(ldats_paper_cpt)
```

## Running LDATS changepoint models

### LDATS LDA and LDATS changepoint

Run the LDATS time series model to identify changepoints on the LDA topics selected by LDATS:

```{r run LDATS LDA and LDATS cpt, eval = params$run_models}
ldats_ldats_results <- TS_on_LDA(LDA_models = ldats_ldamodel,
                                 document_covariate_table = rodents$document_covariate_table,
                                 formulas = ~ sin_year + cos_year,
                                 nchangepoints = 1:6,
                                 timename = "newmoon",
                                 weights = NULL,
                                 control = list(nit = nit))

saveRDS(ldats_ldats_results, file = file.path(vignette_files, "ldats_ldats.RDS"))
```

```{r construct lookup table for LDATS output for changepoint times}
# make the full sequence of possible newmoon values
full_index <- seq(min(rodents$document_covariate_table$newmoon), 
                  max(rodents$document_covariate_table$newmoon))

# generate a lookup table with dates for the newmoons, using `approx` to 
#   linearly interpolate the missing values
ldats_dates <- approx(rodents$document_covariate_table$newmoon, 
                     as.Date(rodents$document_covariate_table$date), 
                     full_index) %>%
  as.data.frame() %>%
  mutate(index = x, 
         date = as.Date(y, origin = "1970-01-01")) %>%
  select(index, date)
```

Select the best time series model and extract the dates of the changepoints:

```{r compute changepoints for LDATS LDA and LDATS cpt}
ldats_ldats_results <- readRDS(file.path(vignette_files, "ldats_ldats.RDS"))

ldats_ldats_cpt <- select_TS(ldats_ldats_results)

ldats_ldats_cpt_dates <- ldats_ldats_cpt$rhos %>%
  as.data.frame() %>%
  reshape::melt() %>%
  dplyr::left_join(ldats_dates, by = c("value" = "index"))
```
